Unshuffling Permutations

### 2015-06-29 article.tex
Partie algebre (suite)

As any coproduct, $\Delta$ leads to the definition of a product obtained
by duality in the following way. From~\eqref{equ:unshuffling_coproduct},
for any permutation $\pi$, we have
\begin{equation}
    \Delta(\pi) =
    \sum_{\sigma, \nu \in S} \lambda_{\sigma, \nu}^{(\pi)} \;
    \sigma \otimes \nu,
\end{equation}
where the $\lambda_{\sigma, \nu}^{(\pi)}$ are nonnegative integers. Now,
by definition of duality, the dual product of $\Delta$ denoted by
$\SHUFFLE$ is a linear binary product on $\QQ[S]$. It satisfies, for any
permutations
$\sigma$ and $\nu$,
\begin{equation}
    \sigma \SHUFFLE \nu =
    \sum_{\pi \in S}
    \lambda_{\sigma, \nu}^{(\pi)} \; \pi.
\end{equation}
We call $\SHUFFLE$ the {\em shuffle product of permutations}. For
instance,
\begin{equation}\begin{split} \label{equ:example_shuffle_product}
    12 \SHUFFLE 21 & =
    1243 + 1324 + 2.1342 + 2.1423 + 3.1432 + 2134 + 2.2314 \\
    & + 3.2341 + 2413 + 2.2431 + 2.3124 + 3142 + 3.3214 + 2.3241 \\
    & + 3421 + 3.4123 + 2.4132 + 2.4213 + 4231 + 4312.
\end{split}\end{equation}
Observe that the coefficient $3$ of the permutation $1432$
in~\eqref{equ:example_shuffle_product} comes from the fact that the
coefficient of the tensor $12 \otimes 21$ is $3$
in~\eqref{equ:example_unshuffling_coproduct}.
\medskip

We say that a permutation $\pi$ {\em appears} in the shuffle
$\sigma \SHUFFLE \nu$ of two permutations $\sigma$ and $\nu$ if the
coefficient $\lambda_{\sigma, \nu}^{(\pi)}$ is different from zero.
Besides, a permutation $\pi$ is a {\em square} if there is a permutation
$\sigma$ such that $\pi$ appears in $\sigma \SHUFFLE \sigma$.
Equivalently, $\pi$ is a square if in the expansion of $\Delta(\pi)$,
there is a tensor $\sigma \otimes \sigma$ with a nonzero coefficient.
\medskip

@@

\begin{proposition}
    Let $\pi$ be a square permutation. Then, $\pi^\sim$, $\bar \pi$,
    and $\pi^{-1}$ are square permutations.
\end{proposition}
\begin{proof}

>>>>

As any coproduct, $\Delta$ leads to the definition of a product obtained
by duality in the following way. From~\eqref{equ:unshuffling_coproduct},
for any permutation $\pi$, we have
\begin{equation}
    \Delta(\pi) =
    \sum_{\sigma, \nu \in S} \lambda_{\sigma, \nu}^\pi \;
    \sigma \otimes \nu,
\end{equation}
where the $\lambda_{\sigma, \nu}^\pi$ are nonnegative integers. Now,
by definition of duality, the dual product of $\Delta$ denoted by
$\SHUFFLE$ is a linear binary product on $\QQ[S]$. It satisfies, for any
permutations
$\sigma$ and $\nu$,
\begin{equation}
    \sigma \SHUFFLE \nu =
    \sum_{\pi \in S}
    \lambda_{\sigma, \nu}^\pi \; \pi.
\end{equation}
We call $\SHUFFLE$ the {\em shuffle product of permutations}. For
instance,
\begin{equation}\begin{split} \label{equ:example_shuffle_product}
    12 \SHUFFLE 21 & =
    1243 + 1324 + 2.1342 + 2.1423 + 3.1432 + 2134 + 2.2314 \\
    & + 3.2341 + 2413 + 2.2431 + 2.3124 + 3142 + 3.3214 + 2.3241 \\
    & + 3421 + 3.4123 + 2.4132 + 2.4213 + 4231 + 4312.
\end{split}\end{equation}
Observe that the coefficient $3$ of the permutation $1432$
in~\eqref{equ:example_shuffle_product} comes from the fact that the
coefficient of the tensor $12 \otimes 21$ is $3$
in~\eqref{equ:example_unshuffling_coproduct}.
\medskip

We say that a permutation $\pi$ {\em appears} in the shuffle
$\sigma \SHUFFLE \nu$ of two permutations $\sigma$ and $\nu$ if the
coefficient $\lambda_{\sigma, \nu}^\pi$ is different from zero.
Besides, a permutation $\pi$ is a {\em square} if there is a permutation
$\sigma$ such that $\pi$ appears in $\sigma \SHUFFLE \sigma$. In this
case, we say that $\sigma$ is a {\em square root} of $\pi$. Equivalently,
$\pi$ is a square if in the expansion of $\Delta(\pi)$, there is a
tensor $\sigma \otimes \sigma$ with a nonzero coefficient.
\medskip

@@

\begin{proposition}
    Let $\pi$ be a square permutation and $\rho$ a square root of $\pi$.
    Then,
    \begin{enumerate}
        \item the permutation $\pi^\sim$ is a square and $\rho^\sim$
        is one of its square roots;
        \item the permutation $\bar \pi$ is a square and $\bar \rho$
        is one of its square roots;
        \item the permutation $\pi^{-1}$ is a square and $\rho^{-1}$
        is one of its square roots.
    \end{enumerate}
\end{proposition}
\begin{proof}
